

This chapter introduces \seq - a sequential implementation
of \algo. The implementation is given first. Then
its work and depth complexities - as introduced in section \ref{section:parmeasures}
- are given.


\section{Implementation}
  \seq is a direct implemenation of the description of \algo
  of section \ref{section:hbalanceintro}.
  First the data structures used are presented.
  Then each component of the implementation is given. Afterwards
  the components are assembled to \seq.
  
  \paragraph{Data Structures}
    The implementation uses two data types.
    \begin{lstlisting}
type Image  = PtrVector (PtrVector Int)
type Hist a = TreeMap Int a
    \end{lstlisting}
    \c{PtrVector a} is a pointer-based array holding values
    of type \c{a}. The use of pointers enables them to be nested.
    Thus they can be directly used to represent two-dimensional images.
    \c{TreeMap k a} is a binary search tree indexed by keys of type
    \c{k} and containing values of type \c{a}. They are
    used for the representation of a histogram.
    \footnote{The use of arrays for the representation of
    histograms has undesireable effects. Languges with
    mutability can destructively update the array
    for each pixel in the image. The Immutability
    of Haskell prevents this and forces the use of
    immutable and shared data-structures like \c{TreeMap k a}.}
    Functions over \c{PtrVector} are suffixed \c{-V}. Functions over \c{TreeMap} aresuffixed \c{-M}.
  
  \paragraph{Histogram Calculation}
    The steps for the creation of the initial histogram are given below:
    \begin{lstlisting}
hist :: Image -> Hist Int
hist = foldrV (\i -> insertWithM (+) i 1) emptyM . concatV
    \end{lstlisting}
    \c{hist} proceeds in two steps. First the image is flatten
    into an one-dimensional array. Then, a \c{TreeMap} is created
    counting the number of occurrences of each gray tone.
    \c{foldrV} is linear in the size of the flattened iamge array.
    \c{insertWithM} is logarithmic in the number elements
    inserted into the map. The size is bounded by the number of 
    gray tones - namely \c{gmax+1}.
    
  \paragraph{Accumulation}
  Calculating the accumulated histogram can be implented by
  a prefix sum over values. It is linear in the size of the map.
  \begin{lstlisting}
accu :: Hist Int -> Hist Int
accu = scanlM (+) 0
  \end{lstlisting}
    
  \paragraph{Normalisation}
  After accumulation, one has to normalise the histogram.
  The normalisation is a direct implementation of its formula in
  \ref{section:hbalanceintro}.
  \begin{lstlisting}
normalize :: Int -> Int -> Hist Int -> Hist Double
normalize a0' agmax' as =
    let a0 = fromIntegral a0'
        agmax = fromIntegral agmax'
        divisor = agmax - a0
    in  mapM (\freq' -> (fromIntegral freq' - a0) / divisor) as
  \end{lstlisting}
  It applies the mapping over the tree-mapusing \c{mapM}.
  \c{fromIntegral} explicitly converts from \c{Int} to \c{Double} since
  Haskell clearly distinguishes them. Variable names with a prime
  (') denote values of type \c{Int}. Variable names without a prime
  denote \c{Double}s. This naming convention is equally used in
  \man, \ndpn and \ndpv.
    
  \paragraph{Scaling}
  Scaling occurs similar to normalisation. It is implemented
  by a mapping over all values in the histogram.
  \begin{lstlisting}
scale :: Int -> Hist Double -> Hist Int
scale gmax = mapM (\d -> floor (d * fromIntegral gmax))
  \end{lstlisting}
    
  \paragraph{Apply}
  The application of the gray tone mapping to the images pixels
  is implemented by a nested \c{mapV} over the image. It uses
  \c{lookupLessEqualM} to lookup the values for the histogram.
  (It reverts back to a lower gray tone, if the gray tone is not found in the map.)
  \begin{lstlisting}
apply :: Hist Int -> Image -> Image
apply as img = mapV (mapV (lookupLessEqualM as)) img
  \end{lstlisting}
  
  \paragraph{\algo}
  Having defined the components, one can now directly define \seq:
  \begin{lstlisting}
hbalance :: Image -> Image
hbalance img =
  let h = hist img
      a = accu h
      a0 = firstM a
      agmax = lastM a
      n = normalize a0 agmax a
      gs = scale gmax n
      img' = apply gs img
  in img'
  \end{lstlisting}
  First the histogram is created (line 3). Then it is accumulated (line 4).
  After that it is normalised (line 5 to 7) and scaled(line 8).
  And finally, the gray tone mapping is applied and returned (line 9 to 10).
  It is defined exactly as previously envisioned.
  
\section{Complexities}
  For sequential algorithms, work and depth fall
  together with their time complexity class.
  This is true for the purely sequential
  program \seq.
  
  Before the time complexity is calculated, 
  a few variables are introduced:
  \begin{itemize}
    \item[n:] the number of pixels in the image
    \item[$h$:] the height of the image and equally the
              the number of subarrays in the nested array
    \item[$w$:] the width of the image and
              equally the length of the subarrays
  \end{itemize}
  
  \begin{table}[h!]
    \caption{Complexities in \seq}
    \label{table:seq}
    \centering
    \begin{tabular}{lll}
        \toprule
        function or variable & $O(...)$ \\
        \midrule
        hbalance          & $n \log gmax + gmax$ \\
        firstM           & 1\\ 
        lastM            & 1\\
        \midrule
        hist              & $n \log gmax$\\
        concatV          & n \\ 
        emptyM           & 1\\ 
        insertWithM      & $\log gmax$\\ 
        foldrV           & $n \log gmax$ \\ 
        \midrule
        accu              & gmax\\ 
        scanlM           & gmax\\
        \midrule
        normalize         & gmax\\ 
        scale             & gmax\\ 
        mapM             & gmax\\ 
        \midrule
        apply             & $n \log gmax = w \cdot h \cdot \log gmax$ \\
        lookupLessEqualM & $\log gmax$ \\ 
    \end{tabular}
  \end{table}
  
  The complexity for the functions involved in \seq are given
  in table \ref{table:seq}.
  The functions are grouped together by their context.
  Each component of \seq (e.g. \c{hist}, \c{accu}) is
  given a group with its each of its subfunctions and
  the components complexity.
  For example, \c{hist} uses three functions.
  \c{concatV} has linear complexity in the number
  of total elements in the nested array - that is \c{n}.
  \c{insertWithM} is an logarithmic time insertion operation
  into a tree. \c{foldrV} traverses each pixel,
  starting with an empty map \c{emptyM} and
  adds each pixel into the TreeMap using \c{insertWithM}.
  There are \c{n} pixels in the array and and each insertion
  takes at-most time logarithmic to the maximum gray tone \c{gmax}.
  Therefore its time complexity is $O(n \log gmax)$.
  Further analysis of the other functions reveals the complexities
  given in the table.
  
  
  Finally, \c{hbalance} uses
  \c{hist} ($O(n \log gmax)$) and a few
  other functions ($O(gmax)$). 
  Therefore \seq has a complexity of $O(n \log gmax + gmax)$.
  
  \p
  The next chapter introduces the first parallel implementation \man.
  
